---
title: "Analyze the ToothGrowth data in the R datasets package"
author: "ِAbdallah Mohammed"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

# summary of the data:

The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC).

## Form:   
**`A data frame with 60 observations on 3 variables.`**  
1. len -> numeric	Tooth length  
2. supp -> factor	Supplement type (VC or OJ).  
3. dose -> numeric	Dose in milligrams/day  

## Source:
C. I. Bliss (1952). The Statistics of Bioassay. Academic Press.

## References: 
* McNeil, D. R. (1977). Interactive Data Analysis. New York: Wiley.

* Crampton, E. W. (1947). The growth of the odontoblast of the incisor teeth as a criterion of vitamin C intake of the guinea pig. The Journal of Nutrition, 33(5), 491–504. doi:10.1093/jn/33.5.491.


# Basic Exploratory Data Analysis:

```{r, echo=TRUE, message=FALSE}
# Load the dataset
library(dplyr)
data("ToothGrowth")

# Convert dose to factor
ToothGrowth$dose <- factor(ToothGrowth$dose)

# Basic structure and summary of the data
str(ToothGrowth)
summary(ToothGrowth)

# Plotting the relationship between length and the two factors (supp and dose)
par(mfrow=c(1,2))
with(ToothGrowth, plot(len ~ supp, main = "Tooth Length by Supplement", 
                       ylab = "Tooth Length", xlab = "Supplement Type"))
with(ToothGrowth, plot(len ~ dose, main = "Tooth Length by Dose", 
                       ylab = "Tooth Length", xlab = "Dose"))
```


# Hypotheses
## Comparison of Supplement Type (VC vs OJ):
* Null Hypothesis (H0): There is no difference in tooth length between the VC supplement and orange juice (OJ).
* Alternative Hypothesis (H1): There is a difference in tooth length between the VC supplement and orange juice (OJ).

## Comparison of Doses:
* Null Hypothesis (H0): There is no difference in tooth length among the different doses of Vitamin C (0.5, 1, and 2 mg/day).
* Alternative Hypothesis (H1): There is a difference in tooth length among the different doses of Vitamin C

# Hypothesis Tests and Confidence Intervals:

In this section, we will use t-tests to compare the tooth length by supplement type (supp) and by dose (dose). The t-tests will help us determine if there are statistically significant differences between groups.


## Comparison by Supplement (VC vs OJ):
```{r}
# T-test to compare tooth length by supplement
p_value_supp <- with(ToothGrowth, t.test(len ~ supp, alternative = "greater"))$p.value
p_value_supp
```
p-value: `r round(p_value_supp, 4)`

## Comparison by Dose:

For the dose comparison, we'll perform pairwise t-tests between the three dose levels (0.5, 1, and 2 mg/day) to see if higher doses result in more tooth growth.

```{r}
# Pairwise t-tests for dose levels
p_value_dose_1_0.5 <- t.test(len ~ dose, data = ToothGrowth %>% filter(dose %in% c(0.5, 1)), 
                             alternative = "less")$p.value
p_value_dose_2_0.5 <- t.test(len ~ dose, data = ToothGrowth %>% filter(dose %in% c(0.5, 2)), 
                             alternative = "less")$p.value
p_value_dose_2_1 <- t.test(len ~ dose, data = ToothGrowth %>% filter(dose %in% c(1, 2)), 
                           alternative = "less")$p.value

p_value_dose_1_0.5
p_value_dose_2_0.5
p_value_dose_2_1
```

p-values:

* p-value for dose 1 vs 0.5: r round(p_value_dose_1_0.5, 4)
* p-value for dose 2 vs 0.5: r round(p_value_dose_2_0.5, 4)
* p-value for dose 2 vs 1: r round(p_value_dose_2_1, 4)


# Conclusions and Assumptions:


## Conclusions:
### Comparison of Supplement Type (VC vs OJ):

* p-value: 0.0303
* Conclusion: Since the p-value is less than 0.05, we reject the null hypothesis (which states that there is no difference). Therefore, we conclude that there is a statistically significant difference in tooth length between the VC supplement and orange juice (OJ). This indicates that the VC supplement has a greater effect on tooth growth compared to orange juice.

### Comparison of Doses:

#### Dose 1 vs Dose 0.5:
* p-value: 6.34e-08
* Conclusion: The p-value is much lower than 0.05, indicating a statistically significant difference between the two doses. We conclude that Dose 1 leads to a statistically significant increase in tooth length compared to Dose 0.5.

#### Dose 2 vs Dose 0.5:
* p-value: 2.20e-14
* Conclusion: The p-value suggests a strong significant difference between the two doses. We conclude that Dose 2 results in a substantial statistically significant increase in tooth length compared to Dose 0.5.

#### Dose 2 vs Dose 1:
* p-value: 9.53e-06
* Conclusion: Again, the p-value is less than 0.05, indicating a statistically significant difference between the two doses. We conclude that Dose 2 leads to a statistically significant increase in tooth length compared to Dose 1.

### Summary:
The results indicate that the use of Vitamin C (either from the VC supplement or orange juice) and higher doses of it positively affect tooth growth in guinea pigs. Overall, we can say that higher doses of Vitamin C significantly enhance tooth growth

## Assumptions:

* The t-test assumes that the data follows a normal distribution.
* Variances between groups are assumed to be equal (in the case of unpaired t-tests).
* Independence of observations is assumed.
